---
title: "R Notebook"
output: github_document 
---


# Libraries
```{r}
library(tidyverse)
library(bnlearn)
library(lattice)
library(corrplot)
```

# Data
```{r}
data <- read_csv("data/sachs1.csv")
col_names <- c('Raf', 'Mek', 'Plcg', 'PIP2', 'PIP3', "Erk", "Akt",
                   "PKA", "PKC", "P38", "Jnk")
names(data) <- col_names
head(data)
glimpse(data)
sum(is.na(data))
nrow(data)
```
# Visualizations
The data is continuous, float, has no missing values. Let's do some visualizations.
```{r}
df_long <- data %>% gather(key = "molecules", value = "concentration")

ggplot(df_long, aes(molecules, concentration)) +
  geom_boxplot()
```

There are some extreme conecntration values for certain molcules. Let's filter for concentration less than 1000. 

```{r}
df_test <- df_long %>% 
  filter(concentration < 1000)

ggplot(df_test, aes(molecules, concentration)) +
  geom_boxplot()
```
PKA seems to be the dominant molecule. Should check what these mean 

```{r}
summary(data)
```
## Concentration distributions

```{r}
ggplot(data, aes(pmek)) +
  geom_histogram(aes(y = .nrow = nrow = ins = 50, color = "white") +
  geom_density()
```

```{r}

# create a histogram for each column and arrange them in a grid
ggplot(df_long, aes(x = concentration)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.8, color = "lightblue") +
  geom_density(alpha = .2) +
  labs(x = "Value", y = "Density") +
  facet_wrap(~molecules, nrow = 4,  ncol = 3, scales = "free")

ggsave("paper/images/molecule_skews.pdf",
  device = pdf
)
```
The distributions of the molecule concentrations are heavily skewed. Most of them are small and cluster around zero. The variables are not symmetric and clearly violate the distributional assumptions underlying GBNs.
## Dependence relationships

```{r}

cor_matrix <- cor(data)
# Customize the correlation plot
pdf("paper/images/molecule_corr.pdf")
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

dev.off()
```

```{r}
ggplot(data, aes(x = PKC, y = PKA)) +
  geom_point() +
  geom_smooth(method = lm)
```
```{r}
ggplot(data, aes(x = PKC, y = PIP3)) +
  geom_point() +
  geom_smooth(method = lm)
```
Looks like the dependence relationships in the data are not always linear. Most conditional independence tests and network scores are designed to capture linear relationships have very low poer in detecting nonlinear ones. Structure learning algorithms using such statistics are unable to correctly learn the arcs in the DAG.

# Charts for paper
```{r}
data <- read_csv("data/sachs1.csv")
ggplot(data, aes(x = PKA)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "#66B2FF",
                 color = "black",
                 bins = 30,
                 size = 0.2) +
  geom_density(size = 0.3) +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(color = "black", fill = NA))

```

```{r}
ggplot(data, aes(x = praf, y = PKA)) +
  geom_point(color = "#66B2FF",
             fill = "black",
             alpha = 0.5) +
  geom_smooth(method = lm,
              color = "black",
              size = 0.4) +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(color = "black", fill = NA))
```

